Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package

### 2020-06-04 docs/article_bigsimr.tex
split up document into sections. did spelling check

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={GIVE ME A TITLE},
  pdfauthor={A.G. Schissler; A.B. PERSON},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage[]{natbib}
\bibliographystyle{apalike}

\title{GIVE ME A TITLE}
\author{A.G. Schissler \and A.B. PERSON}
\date{Monday, June 01, 2020}

\begin{document}
\maketitle

{
\setcounter{tocdepth}{3}
\tableofcontents
}
\hypertarget{prerequisites}{%
\section{Prerequisites}\label{prerequisites}}

This is a \emph{sample} book written in \textbf{Markdown}. You can use anything that Pandoc's Markdown supports, e.g., a math equation \(a^2 + b^2 = c^2\).

The \textbf{bookdown} package can be installed from CRAN or Github:

Remember each Rmd file contains one and only one chapter, and a chapter is defined by the first-level heading \texttt{\#}.

To compile this example to PDF, you need XeLaTeX. You are recommended to install TinyTeX (which includes XeLaTeX): \url{https://yihui.name/tinytex/}.

  \bibliography{bibliography.bib,packages.bib}

>>>>

% Options for packages loaded elsewhere
\PassOptionsToPackage{unicode}{hyperref}
\PassOptionsToPackage{hyphens}{url}
%
\documentclass[
]{article}
\usepackage{lmodern}
\usepackage{amssymb,amsmath}
\usepackage{ifxetex,ifluatex}
\ifnum 0\ifxetex 1\fi\ifluatex 1\fi=0 % if pdftex
  \usepackage[T1]{fontenc}
  \usepackage[utf8]{inputenc}
  \usepackage{textcomp} % provide euro and other symbols
\else % if luatex or xetex
  \usepackage{unicode-math}
  \defaultfontfeatures{Scale=MatchLowercase}
  \defaultfontfeatures[\rmfamily]{Ligatures=TeX,Scale=1}
\fi
% Use upquote if available, for straight quotes in verbatim environments
\IfFileExists{upquote.sty}{\usepackage{upquote}}{}
\IfFileExists{microtype.sty}{% use microtype if available
  \usepackage[]{microtype}
  \UseMicrotypeSet[protrusion]{basicmath} % disable protrusion for tt fonts
}{}
\makeatletter
\@ifundefined{KOMAClassName}{% if non-KOMA class
  \IfFileExists{parskip.sty}{%
    \usepackage{parskip}
  }{% else
    \setlength{\parindent}{0pt}
    \setlength{\parskip}{6pt plus 2pt minus 1pt}}
}{% if KOMA class
  \KOMAoptions{parskip=half}}
\makeatother
\usepackage{xcolor}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\hypersetup{
  pdftitle={Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package},
  pdfauthor={A.G. Schissler; A. Knudson},
  hidelinks,
  pdfcreator={LaTeX via pandoc}}
\urlstyle{same} % disable monospaced font for URLs
\usepackage{color}
\usepackage{fancyvrb}
\newcommand{\VerbBar}{|}
\newcommand{\VERB}{\Verb[commandchars=\\\{\}]}
\DefineVerbatimEnvironment{Highlighting}{Verbatim}{commandchars=\\\{\}}
% Add ',fontsize=\small' for more characters per line
\usepackage{framed}
\definecolor{shadecolor}{RGB}{248,248,248}
\newenvironment{Shaded}{\begin{snugshade}}{\end{snugshade}}
\newcommand{\AlertTok}[1]{\textcolor[rgb]{0.94,0.16,0.16}{#1}}
\newcommand{\AnnotationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\AttributeTok}[1]{\textcolor[rgb]{0.77,0.63,0.00}{#1}}
\newcommand{\BaseNTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\BuiltInTok}[1]{#1}
\newcommand{\CharTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\CommentTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\CommentVarTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ConstantTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ControlFlowTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\DataTypeTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{#1}}
\newcommand{\DecValTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\DocumentationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\ErrorTok}[1]{\textcolor[rgb]{0.64,0.00,0.00}{\textbf{#1}}}
\newcommand{\ExtensionTok}[1]{#1}
\newcommand{\FloatTok}[1]{\textcolor[rgb]{0.00,0.00,0.81}{#1}}
\newcommand{\FunctionTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\ImportTok}[1]{#1}
\newcommand{\InformationTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\newcommand{\KeywordTok}[1]{\textcolor[rgb]{0.13,0.29,0.53}{\textbf{#1}}}
\newcommand{\NormalTok}[1]{#1}
\newcommand{\OperatorTok}[1]{\textcolor[rgb]{0.81,0.36,0.00}{\textbf{#1}}}
\newcommand{\OtherTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{#1}}
\newcommand{\PreprocessorTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textit{#1}}}
\newcommand{\RegionMarkerTok}[1]{#1}
\newcommand{\SpecialCharTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\SpecialStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\StringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\VariableTok}[1]{\textcolor[rgb]{0.00,0.00,0.00}{#1}}
\newcommand{\VerbatimStringTok}[1]{\textcolor[rgb]{0.31,0.60,0.02}{#1}}
\newcommand{\WarningTok}[1]{\textcolor[rgb]{0.56,0.35,0.01}{\textbf{\textit{#1}}}}
\usepackage{longtable,booktabs}
% Correct order of tables after \paragraph or \subparagraph
\usepackage{etoolbox}
\makeatletter
\patchcmd\longtable{\par}{\if@noskipsec\mbox{}\fi\par}{}{}
\makeatother
% Allow footnotes in longtable head/foot
\IfFileExists{footnotehyper.sty}{\usepackage{footnotehyper}}{\usepackage{footnote}}
\makesavenoteenv{longtable}
\usepackage{graphicx,grffile}
\makeatletter
\def\maxwidth{\ifdim\Gin@nat@width>\linewidth\linewidth\else\Gin@nat@width\fi}
\def\maxheight{\ifdim\Gin@nat@height>\textheight\textheight\else\Gin@nat@height\fi}
\makeatother
% Scale images if necessary, so that they will not overflow the page
% margins by default, and it is still possible to overwrite the defaults
% using explicit options in \includegraphics[width, height, ...]{}
\setkeys{Gin}{width=\maxwidth,height=\maxheight,keepaspectratio}
% Set default figure placement to htbp
\makeatletter
\def\fps@figure{htbp}
\makeatother
\setlength{\emergencystretch}{3em} % prevent overfull lines
\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}
\setcounter{secnumdepth}{5}
\usepackage{booktabs}
\usepackage{amsthm}
\makeatletter
\def\thm@space@setup{%
  \thm@preskip=8pt plus 2pt minus 4pt
  \thm@postskip=\thm@preskip
}
\makeatother
\usepackage[]{natbib}
\bibliographystyle{apalike}

\title{Simulating Ultra High-Dimensional Multivariate Data Using the bigsimr R Package}
\author{A.G. Schissler \and A. Knudson}
\date{Thursday, June 04, 2020}

\begin{document}
\maketitle
\begin{abstract}
In this era of Big Data, it is critical to realistically simulate data to conduct informative Monte Carlo studies. This is problematic when data are inherently multivariate while at the same time are (ultra-) high dimensional. This situation appears frequently in observational data found on online and in high-throughput biomedical experiments (e.g., RNA-sequencing). Due to the difficulty in simulating realistic correlated data points, researchers often resort to simulation designs that posit independence --- greatly diminishing the insight into the empirical operating characteristics of any proposed methodology. Major challenges lie in the computational complexity involved in simulating these massive random vectors. We propose a fairly general, scalable procedure to simulate high-dimensional multivariate distributions with pre-specified marginal characteristics and dependency characteristics. As a motivating example, we use our methodology to study large-scale statistical inferential procedures applied to cancer-related RNA-sequencing data sets. The proposed algorithm is implemented as the \texttt{bigsimr} R package.
\end{abstract}

{
\setcounter{tocdepth}{3}
\tableofcontents
}
\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

Massive high-dimensional data sets are commonplace in many areas of scientific inquiry. As new methods are developed for these data, a fundamental challenge lies in designing and conducting simulation studies to assess operating characteristics of proposed methodology, such as false positive rates, statistical power, and robustness --- often in comparison to existing methods. Another application lies in the discovering the sampling distribution of test statistic under hypothesized null models in complex scenarios. Such Monte Carlo studies in the high-dimensional setting often become difficult to conduct with existing algorithms and tools, or require expertise in computing on clusters. This is particularly true when simulating massive multivariate, non-normal distributions.

Along those lines, simulating high-dimension dependent negative binomial data motivates this work --- in pursuit of modeling RNA-sequencing data \citep{Wang2009b, Conesa2016b} derived from breast cancer patients. This laboratory procedure results in count data with infinite support, since RNA-sequencing platforms measure gene expression by enumerating the number of reads aligned to genomic regions. Often researchers posit a negative binomial model (refs) as counts are often over (or under) expressed that a Poisson model would suggest. Yet due to inherent biological processes, gene expression data exhibits correlation (coexpression) across genes \citep{BE07, Schissler2018}. RNA-sequencing analysis has garnered much interest in the statistical literature (refs). Often researchers simulate independently or from their proposed model in order to conduct Monte Carlo studies (refs). An alternative strategy is to think \emph{generatively} about the data collection process and scientific domain knowledge and design simulations with probabilistic models that reflect those processes. In our motivating example, we'll study the consequences correlation in the simulation study replicating test statistics from an illustrative, classical large-scale hypothesis testing study \citep[see][]{Efron2004}.

\hypertarget{ultra-high-dimensional-multivariate-modeling-and-simulation-desired-properties}{%
\subsection{Ultra High Dimensional multivariate modeling and simulation desired properties}\label{ultra-high-dimensional-multivariate-modeling-and-simulation-desired-properties}}

With our application in mind and seeking a general-purpose algorithm with broad applications, our purposed methodology should possess the following properties (adapted from criteria from \citet{Nik13a}):

\begin{itemize}
\tightlist
\item
  P1: Wide range of dependence, allowing both positive and negative dependence
\item
  P2: Flexible dependence, meaning that the number of bivariate marginals is (approximately) equal to the number of dependence parameters.
\item
  P3: Flexible marginal modeling, generating heterogeneous data --- possibly from differing probability families.
\end{itemize}

Moreover, the simulation method must scale to high dimensions:

\begin{itemize}
\tightlist
\item
  S1: Procedure must scale to high dimensions, computable in a reasonable amount time.
\item
  S2: Procedure must scale to high dimensions while maintaining accuracy.
\end{itemize}

As others \citet{MB13} have noted, however, simulating dependent data can be challenging. A central issue lies characterizing dependency between components in the high-dimensional random vector. The choice of correlation in typical practice usually relates to the eventual analytic goal and distributional assumptions of the data (e.g.~non-normal, discrete, finite support, etc). Here we propose a scheme that ignores the analytic goal --- and, instead, aims to match the empirically-derived estimated correlation data and marginal characteristics. For normal data, the Pearson product-moment correlation describes the dependency completely. As we will see, however, simulating arbitrary random vectors with a given Pearson correlation matrix is computationally intense \citep[\citet{Xia17}]{Chen2001}, violating property \textbf{S1}. On the other hand, an analyst may consider the use of non-parametric correlation measures, such as Spearman's \(\rho\) and Kendall's \(\tau\), particularly when modeling non-normal data as in our application. Adding to the complexity is the well-known, marginal-dependent constraints on the possible bivariate correlation (a consequence of the bounds of bivariate distribution functions --- the \emph{Frechet-Hoeffding bounds} \citep{Nelsen2007}). Denote the pointwise upper bound as \(M(y_i, y_{i^\prime})\) These bounds give strict restraints on the possible bounds and cannot be overcome through algorithm design. Yet not all multivariate simulation approaches obtain these bounds {[}for example REF{]}. Computationally, algorithms that rely on serial computation scale poorly to high dimensions (refs - hierarchy) and fail to make use of modern high-performance computing including parallelization and graphical processing unit acceleration \citep{Li2019gpu}.

To meet the desired properties in the face of the challenging setting, we present a general-purpose, scalable multivariate simulation algorithm. The crux of the method lies in the construction of a Gaussian copula {[}refs{]}. The idea that many multivariate and marginally heterogeneous distribution can be constructed in such a manner is well known {[}refs{]}. This article's contribution lies in it's application to high-dimensional data through a high-performance implementation (\texttt{bigsimr}) and speed-focused algorithm design. The algorithm design relies on useful properties of non-standard correlation measures, namely Kendall's \(\tau\) and Spearman's \(\rho\). We'll show that quality of simulation does not require the use of the standard Pearson correlation matrix, even for normally distributed data. The purpose of our method is to generate random vectors that match exactly specified any possible Kendall's \(\tau\) or Spearman's \(\rho\), and approximately Pearson product-moment correlation.

The study proceeds with more details on our motivating example in RNA-sequencing breast cancer data. Then we describe and justify our simulation methodology, followed by extensive Monte Carlo studies under various distributional assumptions --- emphasizing normal and non-normal scenarios. In these studies, we evaluate the accuracy and speed of our approach in comparison to other readily available software. After evaluating \emph{in silico}, we revisit our motivating
example by employing our method and compare our simulations to empirically-derived statistics. Finally, we'll make concluding remarks regarding the method's utility and future directions.

\hypertarget{background-and-notation}{%
\section{Background and notation}\label{background-and-notation}}

\hypertarget{gaussian-copulas}{%
\subsection{Gaussian copulas}\label{gaussian-copulas}}

We present an general-purpose, scalable multivariate simulation algorithm. The crux of the method is construction of a Gaussian copula. This idea is well known {[}refs{]}. A copula is a distribution function on \([0,1]^d\) describing a random vector with standard uniform marginals. Moreover, for any random vector \({\bf X}=(X_1, \ldots, X_d)\) with cumulative distribution function (CDF) \(F\) and marginal CDFs \(F_i\) there is a copula function
\(C(u_1, \ldots, u_d)\) so that

\[
F(x_1, \ldots,x_d) = \mathbb P(X_1\leq x_1, \ldots,X_d\leq x_d) = C(F_1(x_1), \ldots, F_d(x_d)), \,\,\, x_i\in \mathbb R, i=1,\ldots,d. 
\]

A Gaussian copula is the case where all marginal CDFs \(F_i\) are the standard normal cdf, \(\Phi\). The Gaussian copula is the one that corresponds to a multivariate normal distribution with standard normal marginal distributions and covariance matrix \({\bf R}\). (Since the marginals are standard normal, this \({\bf R}\) is also the correlation matrix). If \(F_{{\bf R}}\) is the CDF of such multivariate normal distribution, then the corresponding Gaussian copula \(C_{{\bf R}}\) is defined through

\begin{equation}
\label{eq:gauss}
F_{{\bf R}}(x_1, \ldots, x_d) = C_{{\bf R}}(\Phi(x_1), \ldots, \Phi(x_d)),
\end{equation}

where \(\Phi(\cdot)\) is the standard normal CDF. Note that the copula \(C_{{\bf R}}\) is simply the CDF of the random vector \((\Phi(X_1), \ldots, \Phi(X_d))\), where \((X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R})\).

Sklar's Theorem (ref) guarantee's that any random vector with computational feasible inverse CDFs (P4 above) and obtainable correlation matrix (within the Frechet bounds) can be obtained via transformations involving copula functions. Namely, to simulate a random vector \({\bf Y} = (Y_1, \ldots, Y_d)\), where \(Y_i=F_i^{-1}(U_i)\), \(i=1, \ldots, d\), we can construct a Gaussian copula \({\bf U} = (U_1, \ldots, U_d)\) viz \(U_i=\Phi(X_i)\), \(i=1, \ldots, d\). When an \(Y_i\) is discrete, some care must be taken to define \(F_i\). Letting

\begin{equation}
F_{i}^{-1} = inf\{y:F_{i}(y) \geq u \}
\label{eq:inverseCDF}
\end{equation}

\noindent ensures that \(Y_i \sim F_i\).

Simulation of a general multivariate random vector \({\bf T}\) with margins \(F_i\) based on this copula is quite simple. Ensuring that one obtains a certain dependence among the marginal distributions, however, proves challenging in our proposed Gaussian-copula-based scheme. Two core issues: 1) marginal characteristics induce bounds on the possible bivariate correlations and 2) monotone transformations deform the Pearson correlation which no closed form expression exists in general (ref? Chen2001). The Frechet bounds are well-known.

\hypertarget{measures-of-dependency}{%
\subsection{Measures of dependency}\label{measures-of-dependency}}

The population correlation coefficient, commonly called the Pearson (product-moment) correlation coefficient, describes the linear association between two random variables \(X\) and \(Y\) and is given by

\begin{equation}
\label{eq:pearson}
\rho(X,Y) = \frac{E(XY) - E(X)E(Y)}{\left[ var(X)var(Y)\right]^{1/2}} 
\end{equation}

As \citet{MB13} and \citet{MK01} discuss, for a bivariate normal \((X,Y)\) random vector the Pearson correlation adequately describes the dependency between the components. \(\rho\)'s utility in non-normal or non-linear associations is lacking (\citet{MK01}). Rank-based (ordinal) approaches performance better in these settings, such as Spearman's (denoted \(\rho_s\)) and Kendall's \(\tau\). Define

\begin{equation}
\label{eq:spearman}
\rho_s(X,Y) = 3 \left[ P\left[ (X_1 - X_2)(Y_1-Y_3) > 0 \right] - P\left[ (X_1 - X_2)(Y_1-Y_3) < 0 \right] \right]
\end{equation}

\noindent where \((X_1, Y_1) \overset{d}{=} (X,Y), X_2 \overset{d}{=} X, Y_3 \overset{d}{=} Y\) with \(X_2\) and \(Y_3\) are independent of one other and of \((X_1, Y_1)\). For continuous marginals, the measure works well due to zero probability of ties. For discrete marginals, however, one could perform a rescaled version of \(\rho_s\) for random variables \(X,Y\) with pmfs (or pdfs) \(p(x)\) and \(q(y)\), respectively.

\begin{equation}
\label{eq:spearmanRescaled}
\rho_{RS}(X,Y) = \frac{\rho_s(X,Y)}{ \left[ \left[ 1 - \sum_x p(x)^3 \right] \left[ 1 - \sum_y q(y)^3 \right] \right]^{1/2}}
\end{equation}

\noindent Kendall's \(\tau\) is the probability of concordant pairs minus the probability of discordant pairs and is given compactly as

\begin{equation}
  \label{eq:tau}
  \tau(X,Y) = \frac{\rho_s(X,Y)}{ \left[ \left[ 1 - \sum_x p(x)^3 \right] \left[ 1 - \sum_y q(y)^3 \right] \right]^{1/2}}
\end{equation}

\hypertarget{marginal-dependent-bivariate-correlation-bounds}{%
\subsection{Marginal-dependent bivariate correlation bounds}\label{marginal-dependent-bivariate-correlation-bounds}}

The pairwise correlation between two correlated random variables cannot in general obtain the full range of possible values, \([-1,-1]\). The range, called the Frechet(-Hoeffding) bounds, is a well-known function of the marginal distributions and are given by \citep{BF17}:

\begin{equation}
\label{eq:frechet}
\rho^{max} = \rho \left( F^{-1}_1 (U), F^{-1}_2 (U) \right), \quad \rho^{min} = \rho \left( F^{-1}_1 (U), F^{-1}_2 (1 - U) \right)
\end{equation}

\noindent where \(U\) is a uniform random variable in \((0,1)\), and \(F^{-1}_1, F^{-1}_2\) are the inverse cdf of random variables \(X_1\) and \(X_2\), respectively. For discrete random variables, define \(F^{-1}\) as in Equation \eqref{eq:inverseCDF}.

\hypertarget{simulation-algorithm}{%
\section{Simulation algorithm}\label{simulation-algorithm}}

\hypertarget{algorithm-description}{%
\subsection{Algorithm description}\label{algorithm-description}}

This section describes the method for simulating a random vector \(\bf Y\) with \(Y_i\) components for \(i=1,2,\ldots,d\). Each \(Y_i\) has a specified marginal
distribution function \(F_i\) and its inverse. To characterize dependency, every pair \((Y_i, Y_j)\) has either a specified Pearson correlation \eqref{eq:pearson}, (rescaled) Spearman correlation \eqref{eq:spearmanRescaled} , or Kendall's \(\tau\) (Equation \eqref{eq:tau} ). The method only approximately matches the Pearson correlation in general, whereas the rank-based methods are exact.

The method is best understand as a \textbf{parallelized Gaussian copula} (see Equation \eqref{eq:gauss}. We shall see that constructing continuous joint distributions that match a target Spearman or Kendall's correlations computes easily when employing Gaussian copulas, since this measures are invariant under the monotone transformations involved {[}refs{]}. To do this, we take advantage of a closed form relationship {[}ref?{]} between Kendall's \(\tau\) and Pearson's correlation coefficient for bivariate normal random variables:

\begin{equation}
\label{eq:convertKendall}
r_{Pearson} = sin \left( \tau_{Kendall} \times \frac{\pi}{2} \right), 
\end{equation}

\noindent and similarly for Spearman's \(\rho\) \citep{K58},

\begin{equation}
\label{eq:convertSpearman}
\rho_{Pearson} = 2 \times sin \left( \rho_{Spearman} \times \frac{\pi}{6} \right).
\end{equation}

For discrete marginals, achieving a target Spearman correlation under this scheme is possible by using components from Equation \eqref{eq:spearmanRescaled} to further adjust the input correlation matrix. Let the unscaled Spearman correlation coefficients be \(\rho_{s} \left(Y_{i}, Y_{i^\prime}\right)\) for two marginal distributions and divide the target correlation by the product in the denominator of Equation \eqref{eq:spearmanRescaled}. Let these adjustment factors be denoted as \(a_i = \left[ 1 - \sum_y p_i(y)^3 \right]^{1/2}\) and specifically rescale the target Spearman correlation matrix by

\begin{equation}
\label{eq:convertSpearmanDiscrete}
\rho_{rs} \left(Y_{i}, Y_{i^\prime}\right) = \frac{\rho_{s} \left(Y_{i}, Y_{i^\prime}\right)}{a_i \times a_{i^\prime}}.
\end{equation}

In a similar fashion, we rescale Kendall's \(\tau\) to adjust the input correlation matrix. The conversion formula is given by

\begin{equation}
\label{eq:convertKendallDiscrete}
\rho_{rs} \left(Y_{i}, Y_{i^\prime}\right) = \frac{\rho_{s} \left(Y_{i}, Y_{i^\prime}\right)}{a_i \times a_{i^\prime}}.
\end{equation}

In contrast the rank-based correlations, matching specified Pearson correlation coefficients exactly is computational intense in this scheme. In general, there is no closed form correspondence and involving computing or approximating \(\binom{d}{2}\) integrals of the form \(EY_iY_j = \int \int y_i y_j f_{X|r}(F_i^{-1}(\Phi(z_i)), F_j^{-1}(\Phi(z_j))dy_idy_j\), for \(i,j=1,2,\ldots,d\). For accurate numeric approximation of these integrals, the functions must be evaluated hundreds of times. Others have used efficient Monte Carlo integration schemes (see \citet{Chen2001}), but scale poorly to large dimension in reasonable times (property \textbf{S2}). Despite all this, if one does desire to characterize dependency using Pearson correlations, we often see in practice --- and it is theoretically justified under certain conditions (\citet{Song00}) --- that simply using the target Pearson correlation matrix as the initial conditions to our proposed algorithm will lead to approximate matching in the resultant distribution.

\hypertarget{simulation-algorithm-1}{%
\subsection{Simulation Algorithm}\label{simulation-algorithm-1}}

Putting the together the facts provided in the equations above, we come the following proposed simulation algorithm to produce a random vector \({\bf Y}\) with specified Spearman's correlation and marginal distributions. Note that all computational steps can be parallelized as each operation can be done or either the \(d\) marginals or the \(\binom{d}{2}\) pairs for the correlation values. Even the generation of multivariate normal random vectors is parallelized through optimized matrix multiplication/decomposition routines.

\hypertarget{inputs}{%
\subsubsection{Inputs}\label{inputs}}

\begin{enumerate}
\def\labelenumi{(\arabic{enumi})}
\tightlist
\item
  Marginal characteristics including the same of distributional family and parameter values, denotes as \(F_i\) for marginal component random variable \(Y_i\) for \(i=1,\ldots,d\).
\item
  A target (specified) Spearman's correlation matrix \({\bf R_{Spearman}}\) with
  each \(\rho_{Spearman}\) within the Frechet limits. Alternatively, the rescaled
  Spearman's correlation (see Equation \eqref{eq:convertKendallDiscrete} ) can be provided.
\item
  An error tolerance \(\epsilon\) for finding the nearest positive definite matrix a transformed correlation matrix (see step iii below). Alternatively, a maximum number of iterations can be supplied.
\end{enumerate}

\hypertarget{algorithm}{%
\subsubsection{Algorithm}\label{algorithm}}

\begin{enumerate}
\def\labelenumi{(\roman{enumi})}
\tightlist
\item
  Compute Spearman's \(\rho_{rs}\) (see Equation
  \eqref{eq:convertKendallDiscrete}) from the specified \(\rho_{spearman}\) for each
  pair of marginal random variables \((Y_i,Y_{i^\prime})\) with \(i \neq i^\prime\)
  when either marginal is discrete. In such large scale computations, it may be
  that numerically \(\rho_{rs}\) is be larger/smaller than the Frechet bounds (or
  even \(\pm 1\)). To guard against this, set \(\rho_{rs} = min( \rho_{rs}, M)\),
  where \(M\) is the upper Frechet bound (see {[}Section Introduction{]}) and similarly set \(\rho_{rs} = max( \rho_{rs}, W)\), where \(W\) is the lower Frechet bound for this pair of distributions. Gather these \(\rho_{rs} \lq s\) into a new input correlation matrix \({\bf R_{rs}}\).\\
\item
  Convert \({\bf R_{Spearman}}\) into \({\bf R_{Pearson}}\) via \(\rho_{Pearson} = 2 \times sin \left( \rho_{rs} \times \frac{\pi}{6} \right)\).\\
\item
  Finally, ensure that \({\bf R_{Pearson}}\) is positive definite, by finding the nearest positive definite correlation matrix \({\bf R}\) --- using the \texttt{R} routine \texttt{nearPD} in the \texttt{Matrix} package --- within an error tolerance of \(\epsilon\).\\
\item
  Generate \({\bf X}=(X_1, \ldots, X_d) \sim N_d({\bf 0}, {\bf R})\);\\
\item
  Transform \({\bf X}\) to \({\bf U} = (U_1, \ldots, U_d)\) viz \(U_i=\Phi(X_i)\), \(i=1, \ldots, d\);\\
\item
  Return \({\bf Y} = (Y_1, \ldots, Y_d)\), where \(Y_i=F_i^{-1}(U_i)\), \(i=1, \ldots, d\);
\end{enumerate}

\hypertarget{the-bigsimr-r-package}{%
\section{\texorpdfstring{The \texttt{bigsimr} R package}{The bigsimr R package}}\label{the-bigsimr-r-package}}

Short description and key features, including GPU acceleration.

The most computationally extensive lies in the second step of our algorithm. We
use a optimized and parallelized multivariate normal simulator within the \texttt{R}
package \texttt{mvnfast}. The rest of the code is also parallelized, but these steps
run rapidly in serial computation except for extremely large dimension.

\hypertarget{basic-use}{%
\subsection{Basic use}\label{basic-use}}

short description

\hypertarget{specifying-marginals}{%
\subsubsection{Specifying marginals}\label{specifying-marginals}}

As stated earlier, to generate multivariate data, we need a list of marginals (and their parameters), and a correlation structure (matrix). The marginal distributions can be built up as a list of lists, where each sublist contains the information for the target distribution.

The things to point out here are that in each sublist (marginal), the first item is an unnamed character string with the R name of the distribution \emph{without a letter prefix}. E.g. instead of \texttt{rnorm}, we pass in just \texttt{"norm"}. The second thing to note is that the remaining items are \emph{named} arguments that go along with the distribution. A full list of built-in distributions is found in the appendix.

\hypertarget{specify-correlation}{%
\subsubsection{Specify correlation}\label{specify-correlation}}

The next step is to define a correlation structure for the multivariate distribution. This correlation matrix can either come from observed data, or we can set it ourselves, or we can generate a random correlation matrix via \texttt{bigsimr::rcor}. Let's create a simple correlation matrix where all off-diagonal elements are 0.5. Since we have 3 marginals, we need a \(3\times 3\) matrix.

\begin{verbatim}
     [,1] [,2] [,3]
[1,]  1.0  0.5  0.5
[2,]  0.5  1.0  0.5
[3,]  0.5  0.5  1.0
\end{verbatim}

Finally we can generate a random vector with our specified marginals and correlation structure. The last argument, \texttt{type}, is looking to know what kind of correlation matrix it is receiving. Right now it can handle Pearson, Spearman, or Kendall.

\hypertarget{small-sample}{%
\subsubsection{Small sample}\label{small-sample}}

On my machine, there is no dedicated GPU, so I would see the following warning message once per session.

Taking a look at our random vector, we see that it is 10 rows and 3 columns, one column for each marginal.

\begin{verbatim}
       [,1]     [,2] [,3]
 [1,] 3.046 0.003646    2
 [2,] 3.180 0.091149    5
 [3,] 3.070 0.014344    2
 [4,] 3.157 0.164819    2
 [5,] 3.222 0.529193    5
 [6,] 3.118 0.341439    5
 [7,] 3.056 0.028928    2
 [8,] 3.187 0.119534    4
 [9,] 3.118 0.413526    3
[10,] 3.225 0.431735    3
\end{verbatim}

We can simulate many more samples and then check the histogram of each margin, as well as the estimated correlation between the columns.

\hypertarget{scaling-up-n}{%
\subsubsection{Scaling up N}\label{scaling-up-n}}

\begin{center}\includegraphics[width=0.85\linewidth]{article_bigsimr_files/figure-latex/unnamed-chunk-14-1} \end{center}

\hypertarget{evaluation}{%
\subsubsection{Evaluation}\label{evaluation}}

\begin{verbatim}
       [,1]   [,2]   [,3]
[1,] 1.0000 0.4886 0.4945
[2,] 0.4886 1.0000 0.4882
[3,] 0.4945 0.4882 1.0000
\end{verbatim}

We can see that even with 10,000 samples, the estimated correlation of the simulated data is not exactly the same as the target correlation. This can be explained by the fact that some correlations are simply not possible due to the discrete nature of certain distributions. Another possibility is that the copula algorithm is biased and needs correction.

\hypertarget{monte-carlo-evaluation-and-comparsions-to-other-software}{%
\section{Monte Carlo evaluation and Comparsions to other software}\label{monte-carlo-evaluation-and-comparsions-to-other-software}}
